---
title: Covid-19 Comorbidity in Mexico
subtitle: Project for the subject Data Acquistion, Analysis and Scientific Methods in Life Sciences.  
author: 
   - affiliation: Universidad Politécnica de Valencia.
     email: czmeystr@posgrado.upv.es
     name: Cléa Zoe Meystre
     url: https://github.com/Cmey1338

   - affiliation: Universidad Politécnica de Valencia.
     email: knmabalo@posgrado.upv.es
     name: Kenneth Neil Mabalot
     url: https://github.com/neilmabalot-34404 

   - affiliation: Universidad Politécnica de Valencia.
     email: rhgutmor@posgrado.upv.es
     name: Raymundo Humberto Gutiérrez Morán
     url: https://github.com/raymoran-netizen

title-block-banner: true
date: last-modified
date-format: iso
lang: en-US
code-fold: true
code_folding: hide 
toc: true
toc-depth: 6
toc-location: body
format:
  html:
    self_contained: TRUE
    embed-resources: true
editor: 
  markdown: 
    wrap: sentence
---

```{r setup, include=FALSE}

#Load libraries
library(dplyr)
library (kableExtra)
library(knitr)
library(broom)
library(ggplot2)
library(tidyr)

#Dataset
data <- "data/covid_mexico_comorbidity_dataset_100k.csv"
covid19A <- read.table(data, header = T, sep = ",")

```

# Introduction 

  Experiencing a global pandemic in the 21st century is scientifically and statistically jarring, particularly given the era’s sophisticated pharmaceutical and surveillance capabilities.
The emergence of a highly contagious respiratory illness in 2020 infamously known as the Coronavirus Disease 2019 (COVID-19), caused by the virus ‘Severe Acute Respiratory Syndrome Coronavirus 2’ (SARs-CoV-2), has profoundly impacted global health that resulted in a declaration of global health emergency. As a result, this prompted extensive research into the factors that contribute to severe outcomes and mortality.

  Understanding the pathophysiological mechanism of viral disease infections, the role of comorbidities, i.e., pre-existing health conditions, became the main research interest linking its influence on the risk of death among individuals infected with SARS-CoV-2.
Over the years since the pandemic passed, multiple literature quantified the mortality risk associated with comorbidities.

  One study by Elezkurtaj et al. (2021) highlighted that COVID-19 patients with more than two comorbidities increased mortality than those with single or no comorbidities.
Among the comorbidities identified, a study by Espinosa et al. (2021) revealed that hypertension, diabetes, and cardiovascular disease were the most frequent comorbidities in fatal cases of COVID-19.
Furthermore, another study by Salinas-Aguirre et al. (2022) concluded that diabetes mellitus, systemic arterial hypertension, obesity, and chronic kidney damage increase mortality in patients with COVID-19 in the population of Coahuila in Mexico.
These findings were further solidified by a study conducted by Lowhorn et al. (2024) stating that hypertension, followed by obesity and diabetes, are the most often occurring comorbidities in COVID-19 patients and are also some of the comorbidities most strongly related with high mortality rates in the Mexican population.

  Considering the literatures cited, the impact of the COVID-19 pandemic in Mexico is arguably one of the most severe in the world.
The pandemic in Mexico wasn't just a viral outbreak; it was a synergistic epidemic where the virus collided with the country's high rates of pre-existing chronic disease crisis (comorbidities).
In line with this, our project aims to statistically analyze the data from Mexico City presented in the Analytical Viewer for COVID-19, a platform created by Mexico's National Institute of Statistics and Geography (INEGI).
By exploring this link, our project aims to delve deeper into the nuances of comorbidity-related risks and its implications on actionable insights and protective measures for future pandemic preparedness, especially to the vulnerable sector of the population.

# Objective
The current project aims to substantiate the association between the mortality and the pre-existing health conditions / comorbidities of COVID-19 patients from Mexico City using the INEGI dataset, thereby providing a comparative analysis against literature-based evidence from some parts of Mexico and national studies to validate the consistency of comorbidity impact across different Mexican demographics.

# Hypotheses
1. COVID-19 patients with pre-existing health condition / comorbidity are more likely to die from the viral disease.

2. Among the 4 identified comorbidities, COVID-19 patients with hypertension are more likely to die from the viral disease.

3. Ultimately, COVID-19 patients with hypertension in combination of another comorbidity are more likely to die from the viral disease.

# Materials and Methods 
1. Chi square Test - The Chi square test is gonna allow us to find statistical association between deaths & each comorbidity.
All tests resulted in highly significant p-values, indicating that the distribution of deaths differs significantly between patients with and without the comorbidity.
So the death rate for those with any comorbidities is not likely to be random.
This supports the conclusion that comorbidities are strongly associated with increased mortality.
Once statistically significant values are found.
An anylsis to see association between each and every comorbidity would be ideal, to investigate about the probability of death based on the presence or absence of each comorbidity, combinations & interactions.
The logistic model is ideal for this analysis because our outcome variables are binary.

2.  Logistic Regressions (LRM)- We first ran individual LRM to measure the association of each individual Comorbidity.
    Then to isolate the effect we incorporated all the Comorbidities analyzed in this study as a Multivariable Model (MVM).
    This model is considering the whole dataset.
    Answering the question: Across the entire population, does each comorbidity increase the odds of dying?
    We then made another version of the MVM that would only consider the people with at least 1 comorbility (a subset of the data (Number of CM \>=1)) that would attenuate the effect of the Comorbitiy (MVMwithCM).
    Answering the question: Among people who already have comorbidities, which specific diseases increase the odds of death?
    We then compared the MVM and the MVMwithCM models with a mortality baseline given by a LRM of only the people WITH OUT any CM (LRMnoCM).
    Another interesting question to analyse would be if the addition of multiple Comorbities increases the odds of death.To answer this question we ran a LRM considering Number_of_Comorbidities (given in our dataset) (LRM_NCM).
    This confirmed that the risk of death increases as the Number_of_Comorbidities increases.
    Identifying which combination of Comorbidities is the most dangerous would help us shape the profile for a high risk of death infected person, given that usually Comorbidities appear simultaneously.
    To analyse this we used the Paired LRM (PLRM) and used it to compare Diabetes + Hypertension, Diabetes + Smoking, Diabetes + Obesity, Hypertension + Smoking, Obesity + Smoking and Hypertension + Obesity.
    Finally, we created a single combined output table with the Individual LRM results for each Comorbidity and the PLRM results.

3.  AIC and BIC analyses - The purpose of AIC (Akaike Information Criterion) and BIC (Bayesian Information Criterion) is to help select the best statistical model from a set of candidates by balancing model fit (how well it explains the data) with model complexity (number of parameters).
    The model with the lowest AIC or BIC value is generally preferred.
    AIC favors predictive accuracy, while BIC leans towards simpler, more parsimonious models, especially with large datasets, aiming closer to the "true" underlying model.
    In our code we found that out of all Single-Predictor Models, Obesity has the lowest AIC and lowest BIC.
    This means that it is the strongest single predictor of COVID death (statistically, based on these models).
    Hypertension is very close, meaning it is also a strong predictor.
    Smoking has the worst AIC/BIC, meaning it is the weakest single predictor.
    Then when interpreting Interpretation of Paired-Predictor Models the pair Diabetes + Smoking had the lowest AIC and BIC values, meaning that they are the best predicts COVID death and is consequently the strongest.
    The AIC/BIC values help identify which combinations add the most predictive value.

4.  AIV and GIV analyses - Incremental value analysis was used to quantify the relative importance of four predictors—hypertension, obesity, diabetes, and smoking—across multiple logistic regression specifications.
    Average Incremental Value (AIV) was computed for each predictor as the mean absolute standardized coefficient (\|β/SE\|) across all models in which the predictor appeared.
    AIV reflects the average individual contribution of a predictor to model performance, regardless of other covariates.
    General Incremental Value (GIV) was calculated as the sum of AIV contributions across all model combinations involving each predictor.
    GIV reflects the overall importance of each variable across the entire model space, capturing how consistently and strongly it improves predictive performance.
    Both metrics provide a complementary assessment: AIV reflects effect strength within models, while GIV reflects global contribution across models.
    Higher values for either metric indicate stronger incremental predictive importance.

5.  ANOVA - The purpose of ANOVA (Analysis of Variance) is to determine if there are statistically significant differences between the means of three or more groups.
    Furthermore it tests if variations in a dependent variable are due to an independent variable or random chance.
    Our ANOVA testing gave us higher LR Chi-square values for Hypertension and Diabetes.
    This confirms that they are both highly significant independent predictors of COVID-19 death.
    All predictors (Diabetes, Hypertension, Smoking, Obesity) have p-values \< 2.2e-16, meaning extremely strong evidence that each predictor is associated with COVID-19 mortality, independently of the other variable in the model.
    Strength of predictors, based on the LR Chisq value is the strongest for Hypertension, then for Diabetes, then Obesity and lastlt Smoking is slightly less strong than others.

# Results

## Chi square Test

```{r}

##load library 
library(broom)

comorbidities <- c("Diabetes", "Hypertension", "Obesity", "Smoking")

chi_results <- data.frame(
  Comorbidity = character(),
  Chi_square = numeric(),
  df = numeric(),
  P_value = numeric()
)

for (c in comorbidities) {
  tab <- table(covid19A$COVID_Death, covid19A[[c]])
  chi <- chisq.test(tab)
  
  chi_results <- rbind(
  chi_results,
  data.frame(
    Comorbidity = c,
    Chi_square  = as.numeric(chi$statistic),
    df          = as.numeric(chi$parameter),
    P_value     = formatC(chi$p.value, format = "e", digits = 3),
    signif = case_when(
      as.numeric(chi$p.value) < 0.001 ~ "***",
      as.numeric(chi$p.value) < 0.01  ~ "**",
      as.numeric(chi$p.value) < 0.05  ~ "*",
      as.numeric(chi$p.value) < 0.1   ~ ".",
    )
  )
)
}


#chi_results
knitr::kable(chi_results, caption = "Table 1. Summary of Chi Square Tests", align = "c") %>%
kable_styling(full_width = FALSE) %>%
footnote(general = "Significance codes: *** p < 0.001, ** p < 0.01, * p < 0.05, . p < 0.1", general_title = "")

```

  Table 1 showing the results of Chi-square test confirms that the presence of four comorbidities namely Diabetes, Hypertension, Obesity, or Smoking is significantly non-randomly associated with mortality, as shown in the low p-values (p<0.001). Hypertension displays the strongest statistical signal of association supporting its prioritization as a critical risk factor.


## Logistic Regression Model (LRM)

LRM for each Comorbidity Individually

```{r}
##load library 
library(broom)

#Indiv. LRM Diabetes
model_Diabetes <- glm(
  COVID_Death ~ Diabetes,
  data = covid19A,
  family = binomial()
)

#Indiv. LRM Smoking
model_Smoking <- glm(
  COVID_Death ~ Smoking,
  data = covid19A,
  family = binomial()
)

#Indiv. LRM  Obesity
model_Obesity <- glm(
  COVID_Death ~ Obesity,
  data = covid19A,
  family = binomial()
)

#Indiv. LRM  Hypertension
model_Hypertension <- glm(
  COVID_Death ~ Hypertension,
  data = covid19A,
  family = binomial()
)

#Multivariable Model (MVM)
MVM <- glm(COVID_Death ~ Diabetes + Hypertension + Obesity + Smoking,
                  data = covid19A,
                  family = binomial)

MVM_tidy <- tidy(MVM) |>
  filter(term != "(Intercept)") |>
  mutate(
    p.value = formatC(p.value, format = "e", digits = 3),
    signif = case_when(
      as.numeric(p.value) < 0.001 ~ "***",
      as.numeric(p.value) < 0.01  ~ "**",
      as.numeric(p.value) < 0.05  ~ "*",
      as.numeric(p.value) < 0.1   ~ ".",
      TRUE                        ~ ""
    )
  )


#summary(MVM)
knitr::kable(MVM_tidy, caption = "Table 2. Multivariable Model of Individual LRM", digits= 2, align = "c") %>%
kable_styling(full_width = FALSE) %>%
footnote(general = "Significance codes: *** p < 0.001, ** p < 0.01, * p < 0.05, . p < 0.1", general_title = "")

```

Table 2 presents the results of the logistic regression analysis evaluating the individual impact of four key comorbidities—diabetes, hypertension, obesity, and smoking—on the outcome of interest. The data indicates that all four variables are statistically significant positive predictors (p<0.001), with high z-statistics ranging from 16.25 to 19.70. Diabetes demonstrated the strongest association with a coefficient estimate of 0.49, closely followed by hypertension (β=0.48) and smoking (β=0.47), while obesity showed a slightly lower but still highly significant estimate of 0.44. These positive coefficients suggest that the presence of each comorbidity is associated with an increase in the log-odds of the outcome occurring. 


```{r}
#MVM with CM
with_comorb <- subset(covid19A, Number_of_Comorbidities > 0)
MVMwithCM <- glm(COVID_Death ~ Diabetes + Hypertension + Obesity + Smoking,
                  data = with_comorb,
                  family = binomial)

MVMwithCM_tidy <- tidy(MVMwithCM) |>
  filter(term != "(Intercept)") |>
  mutate(
    p.value = formatC(p.value, format = "e", digits = 3),
    signif = case_when(
      as.numeric(p.value) < 0.001 ~ "***",
      as.numeric(p.value) < 0.01  ~ "**",
      as.numeric(p.value) < 0.05  ~ "*",
      as.numeric(p.value) < 0.1   ~ ".",
      TRUE                        ~ ""
    )
  )


#summary(MVMwithCM)
knitr::kable(MVMwithCM_tidy, caption = "Table 3. Multivariable Model with CM of Individual LRM", digits = 2, align = "c")%>%
kable_styling(full_width = FALSE) %>%
footnote(general = "Significance codes: *** p < 0.001, ** p < 0.01, * p < 0.05, . p < 0.1", general_title = "")

```

Table 3 reveals the results of the multivariable logistic regression model incorporating all comorbidities. Consistent with the individual models, all four risk factors: diabetes, hypertension, smoking, and obesity—retained high statistical significance (p<0.001) with z-statistics ranging from 10.92 to 12.47. Diabetes and hypertension demonstrated the strongest associations, both yielding a coefficient estimate of 0.39, while smoking (β=0.38) and obesity (β=0.34) followed with slightly lower estimates. The persistence of these significant positive coefficients suggests that these comorbidities remain robust, independent predictors of the outcome even when adjusted within the multivariable framework.

```{r}
#Logistic Regression 1: Diabetes + Hypertension 
model_Diabetes_Hypertension <- glm( COVID_Death ~ Diabetes + Hypertension, data = covid19A, family = binomial() )

#Logistic Regression 2: Diabetes + Smoking
model_Diabetes_Smoking <- glm( COVID_Death ~ Diabetes + Smoking, data = covid19A, family = binomial() )

#Logistic Regression 3: Diabetes + Obesity
model_Diabetes_Obesity <- glm( COVID_Death ~ Diabetes + Obesity, data = covid19A, family = binomial() )

#Logistic Regression 4: Hypertension + Smoking
model_Hypertension_Smoking <- glm( COVID_Death ~ Hypertension + Smoking, data = covid19A, family = binomial() )

#Logistic Regression 5: Obesity + Smoking
model_Obesity_Smoking <- glm( COVID_Death ~ Obesity + Smoking, data = covid19A, family = binomial() )

#Logistic Regression 6: Hypertension + Obesity
model_Hypertension_Obesity <- glm( COVID_Death ~ Hypertension + Obesity, data = covid19A, family = binomial() )
```

## Individual LRM and Paired LRM

```{r}
#library(dplyr)

models <- list(
  Diabetes_Hypertension = model_Diabetes_Hypertension,
  Diabetes_Smoking = model_Diabetes_Smoking,
  Diabetes_Obesity = model_Diabetes_Obesity,
  Hypertension_Smoking = model_Hypertension_Smoking,
  Obesity_Smoking = model_Obesity_Smoking,
  Hypertension_Obesity = model_Hypertension_Obesity,
  Hypertension = model_Hypertension,
  Obesity = model_Obesity,
  Smoking = model_Smoking,
  Diabetes = model_Diabetes
)

results <- lapply(names(models), function(name) {
  tidy(models[[name]], exponentiate = TRUE, conf.int = TRUE) %>%
    mutate(Model = name)
})
combined_results <- bind_rows(results)
#combined_results

#Single models for Individual CM
models2 <- list(
  Hypertension = model_Hypertension,
  Obesity = model_Obesity,
  Smoking = model_Smoking,
  Diabetes = model_Diabetes
)

results2 <- lapply(names(models2), function(name) {
  tidy(models2[[name]], exponentiate = TRUE, conf.int = TRUE) %>%
    mutate(Model = name)
})
combined_results2 <- bind_rows(results2)
#combined_results2

#Tables of combined results
library(knitr)

combined_results2_clean <- combined_results2 %>%
  filter(term != "(Intercept)") %>%
  select(Model, everything(), -term) %>%
  mutate(
    p.value = formatC(p.value, format = "e", digits = 3),
    signif = case_when(
      as.numeric(p.value) < 0.001 ~ "***",
      as.numeric(p.value) < 0.01  ~ "**",
      as.numeric(p.value) < 0.05  ~ "*",
      as.numeric(p.value) < 0.1   ~ ".",
      TRUE                        ~ ""
    )
  )


knitr::kable(combined_results2_clean, caption = "Table 4. Single-Predictor Logistic Regression Results", digits = 2, align = "c") %>%
kable_styling(full_width = FALSE) %>%
footnote(general = "Significance codes: *** p < 0.001, ** p < 0.01, * p < 0.05, . p < 0.1", general_title = "")

```

Table 4 summarizes the results of the single-predictor logistic regression models, providing the unadjusted odds ratios for each comorbidity. All four risk factors demonstrated highly statistically significant associations with the outcome (p<0.001). Hypertension and diabetes exhibited the strongest magnitude of effect, both yielding an estimate of 1.62, with 95% confidence intervals of 1.55–1.70 and 1.54–1.70, respectively. These findings confirm that, in isolation, each condition serves as a robust predictor of increased odds for the outcome.

```{r}

combined_results_clean <- combined_results %>%
  filter(term != "(Intercept)") %>%
  select(Model, everything(), -term) %>%
  mutate(
    p.value = formatC(p.value, format = "e", digits = 3),
    signif = case_when(
      as.numeric(p.value) < 0.001 ~ "***",
      as.numeric(p.value) < 0.01  ~ "**",
      as.numeric(p.value) < 0.05  ~ "*",
      as.numeric(p.value) < 0.1   ~ ".",
      TRUE                        ~ ""
    )
  )


knitr::kable(combined_results_clean, caption = "Table 5. Paired Logistic Regression Results", digits = 2, align = "c")%>%
kable_styling(full_width = FALSE) %>%
footnote(general = "Significance codes: *** p < 0.001, ** p < 0.01, * p < 0.05, . p < 0.1", general_title = "")

```

Table 5 displays the results of the paired logistic regression analyses, which assess the stability of the associations between the comorbidities and the outcome when adjusting for one another. The data reveals that the risk estimates for diabetes, hypertension, obesity, and smoking remain highly statistically significant (p<0.001) and remarkably consistent across all model configurations. Specifically, diabetes and hypertension consistently showed the strongest associations, with odds ratios (OR) holding steady at approximately 1.62 in both individual and paired models. This stability in effect sizes indicates that these comorbidities function as independent risk factors, as their predictive power is not diminished by the inclusion of concurrent conditions in the model.

## AIC and BIC Analyses

```{r}
models_paired <- list(
  Diabetes_Hypertension = model_Diabetes_Hypertension,
  Diabetes_Smoking      = model_Diabetes_Smoking,
  Diabetes_Obesity      = model_Diabetes_Obesity,
  Hypertension_Smoking  = model_Hypertension_Smoking,
  Obesity_Smoking       = model_Obesity_Smoking,
  Hypertension_Obesity  = model_Hypertension_Obesity
)

info_criteria_paired <- lapply(names(models_paired), function(name) {
  m <- models_paired[[name]]
  
  data.frame(
    Model = name,
    AIC  = AIC(m),
    BIC  = BIC(m)
  )
})

AIC_BIC_table_paired <- dplyr::bind_rows(info_criteria_paired)
#AIC_BIC_table_paired

models2 <- list(
  Hypertension = model_Hypertension,
  Obesity = model_Obesity,
  Smoking = model_Smoking,
  Diabetes = model_Diabetes
)

info_criteria2 <- lapply(names(models2), function(name) {
  m <- models2[[name]]
  
  data.frame(
    Model = name,
    AIC  = AIC(m),
    BIC  = BIC(m)
  )
})

AIC_BIC_table2 <- dplyr::bind_rows(info_criteria2)
#AIC_BIC_table2

knitr::kable(AIC_BIC_table2, caption = "Table 6. AIC and BIC for Single-Predictor Models", digits = 2, align = "c")
```

Table 6 evaluates the relative quality of the single-predictor logistic regression models using the Akaike Information Criterion (AIC) and Bayesian Information Criterion (BIC), where lower values indicate a better trade-off between goodness-of-fit and model complexity. The model utilizing Hypertension as the sole predictor demonstrated the superior statistical fit, yielding the lowest AIC (65,453.78) and BIC (65,472.81) among the four comorbidities analyzed. This suggests that while all predictors are significant, Hypertension provides the most informative individual explanation for the data variance within this set.

```{r}
knitr::kable(AIC_BIC_table_paired, caption = "Table 7. AIC and BIC for Paired Models", digits = 2, align = "c")
```

Table 7 reveals that the model combining Diabetes and Hypertension provided the best fit among all paired combinations, achieving the lowest AIC (65,131.99) and BIC (65,160.53). This was followed by the Hypertension and Obesity model (AIC = 65,147.32) and the Diabetes and Obesity model (AIC = 65,188.86), while the combination involving Smoking generally resulted in higher information criterion values (e.g., Obesity_Smoking AIC = 65,268.05). These results suggest that the synergistic inclusion of Diabetes and Hypertension offers the most parsimonious and effective explanation for the outcome variance among the paired models tested.

## Graphs
Graph a Heatmap to represent results.

```{r}

library(ggplot2)

#Heatmap for Individual CM
#ls(combined_results2)
#head(combined_results2)

#Adding legend and caption to the Heatmap graph
caption_text <- "Figure 1: Statistical significance of risk factors. Darker red = higher significance level (-log10 p-value)."

combined_results2 %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = Model, y = term, fill = -log10(p.value))) +
  geom_tile(color = "grey70") +
  scale_fill_gradient(low = "white", high = "red", name = "-log10(p)") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    plot.caption = element_text(hjust = 0.5)  
  ) +
  labs(
    title = "Individual Comorbidities Risk Factor Heatmap",
    x = "Model",
    y = "Risk Factor",
    caption = strwrap(caption_text, width = 110)
  )

```

Figure 1 presents a heatmap illustrating the statistical significance of the associations between individual comorbidities and the outcome, expressed as -log10(p-value). The color intensity corresponds to the magnitude of the significance, with darker red indicating a more highly significant relationship. The heatmap shows that Hypertension and Diabetes have the highest levels of statistical significance, represented by the darkest red tiles, with -log10(p-values) exceeding 80. Obesity also shows a strong significant association, indicated by a medium-dark red tile corresponding to a -log10(p-value) around 70-75. Smoking appears to have the lowest relative significance among the four factors, represented by a white tile, suggesting its -log10(p-value) is below 60 in this comparative view, although it remains statistically significant based on the tabular data.


```{r}

#ls(combined_results)
#head(combined_results)

#Adding legend and caption to the Heatmap graph
caption_text <- "Figure 2: Statistical significance of risk factors. Darker red = higher significance level (-log10 p-value)." 

combined_results %>%
  filter(term != "(Intercept)") %>%
  ggplot(aes(x = Model, y = term, fill = -log10(p.value))) +
  geom_tile(color = "grey70") +
  scale_fill_gradient(low = "white", high = "red", name = "-log10(p)") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14),
    axis.title = element_text(face = "bold"),
    plot.caption = element_text(hjust = 0.5)  
  ) +
  labs(
    title = "Paired Comorbidities Risk Factor Heatmap",
    x = "Model",
    y = "Risk Factor",
    caption = strwrap(caption_text, width = 110)
  )
```

Figure 2 reveals that a distinct hierarchy in the robustness of the paired associations of each comorbidity. Hypertension exhibits the most profound statistical signal, represented by the highest intensity red tile, followed closely by Diabetes and Obesity, which display strong but progressively lower magnitudes of significance. Notably, the tile corresponding to Smoking appears unpigmented, indicating that while the association is statistically significant (as confirmed in tabular data) the p-value falls below the visualization’s lower threshold of 60. This visualization underscores that while all four factors are predictors, Hypertension and Diabetes provide the highest degree of statistical certainty within the modeled cohort.

## AIV and GIV Analyses

```{r}
# Extract variables from model name
if (is.list(results) && !is.data.frame(results)) {
  results <- bind_rows(results)
}
model_vars <- results %>%
  distinct(Model) %>%
  mutate(vars = strsplit(Model, "_"))

aiv <- results %>%
  filter(term != "(Intercept)") %>%
  mutate(standardized = abs(estimate / std.error)) %>%
  group_by(term) %>%
  summarise(AIV = mean(standardized)) %>%
  arrange(desc(AIV))

# Expand model × variable pairs
library(tidyr)
model_var_pairs <- model_vars %>%
  unnest(vars)%>%
  rename(term = vars)

giv <- model_var_pairs %>%
  left_join(aiv, by = "term") %>%
  group_by(term) %>%
  summarise(GIV = sum(AIV)) %>%
  arrange(desc(GIV))

#Adding legend and caption to the AIV and GIV graph
library(stringr)

caption_text <- "Figure 3: The AIV of the 4 comorbidities are ordered from highest to lowest contribution to COVID-19 mortality."

ggplot(aiv, aes(x = reorder(term, AIV), y = AIV)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Average Incremental Value (AIV) by Predictor",
    x = "Predictor",
    y = "AIV",
    caption = str_wrap(caption_text, width = 70)   # <— caption added + wrapped
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.caption = element_text(
      hjust = 0,                 # left-align caption
      size = 9,
      margin = margin(t = 10)    # spacing above caption
    )
  )
```

Figure 3 displays a horizontal bar plot quantifying the Average Incremental Value (AIV) for four selected comorbid risk factors, ordered by their relative contribution to COVID-19 mortality. The y-axis categorizes the predictors—Hypertension, Obesity, Diabetes, and Smoking—while the x-axis measures their AIV magnitude. The results indicate a hierarchical ranking of importance, with Hypertension exhibiting the highest average incremental value, followed closely by Obesity and Diabetes, both of which surpass an AIV threshold of 60. Smoking demonstrates the lowest relative contribution among the modeled comorbidities, with an AIV falling below 60.

```{r}
giv_caption <- "Figure 4: The GIV of the 4 comorbidities are ordered from highest to lowest contribution to COVID-19 mortality."

ggplot(giv, aes(x = reorder(term, GIV), y = GIV)) +
  geom_col(fill = "darkorange") +
  coord_flip() +
  labs(
    title = "General Incremental Value (GIV) by Predictor",
    x = "Predictor",
    y = "GIV",
    caption = str_wrap(giv_caption, width = 70)   # wrapped caption
  ) +
  theme_minimal(base_size = 10) +
  theme(
    plot.caption = element_text(
      hjust = 0,                 # left align caption
      size = 9,
      margin = margin(t = 10)    # add space above caption
    )
  )
```

Figure 4 shows a horizontal bar plot that illustrates the General Incremental Value (GIV) of four comorbidities in relation to COVID-19 mortality, with the predictors ordered from highest to lowest contribution. The y-axis displays the comorbid predictors—Hypertension, Obesity, Diabetes, and Smoking—while the x-axis quantifies the magnitude of their GIV. The results show a clear hierarchical ranking, where Hypertension demonstrates the largest General Incremental Value, with its bar exceeding the 250 mark on the GIV scale. Obesity and Diabetes follow with similarly high values, both also extending to approximately 250, whereas Smoking exhibits the lowest relative contribution among the tested factors, with a GIV falling between 200 and 250.

## ANOVA Analyses

```{r}
#Comparisons of ANOVA
#anova(model_Diabetes, model_Hypertension, test = "LRT")
#anova(model_Diabetes, model_Smoking, test = "LRT")
#anova(model_Diabetes, model_Obesity, test = "LRT")
#anova(model_Hypertension, model_Smoking, test = "LRT")
#anova(model_Obesity, model_Smoking, test = "LRT")
#anova(model_Hypertension, model_Obesity, test = "LRT")

# Create a data frame with your predictors and model deviances
predictor_summary <- data.frame(
  Predictor = c("Diabetes", "Hypertension", "Obesity", "Smoking"),
  Deviance_vs_baseline = c(
    0,        # Diabetes as baseline
    43.186,   # Hypertension vs Diabetes
    -16.54,   # Obesity vs Diabetes
    -81.759   # Smoking vs Diabetes
  )
)

# Rank predictors by deviance (higher = better predictor)
predictor_summary$Rank <- rank(-predictor_summary$Deviance_vs_baseline, ties.method = "first")

# Sort table by rank
predictor_summary <- predictor_summary[order(predictor_summary$Rank), ]

# Print summary table
#print(predictor_summary)

#library(knitr)

# Create a data frame with your predictors and deviance compared to Diabetes
predictor_summary <- data.frame(
  Predictor = c("Diabetes", "Hypertension", "Obesity", "Smoking"),
  Deviance_vs_Diabetes = c(
    0,        # Diabetes as baseline
    43.186,   # Hypertension vs Diabetes
    -16.54,   # Obesity vs Diabetes
    -81.759   # Smoking vs Diabetes
  )
)

# Rank predictors by deviance (higher = stronger predictor)
predictor_summary$Rank <- rank(-predictor_summary$Deviance_vs_Diabetes, ties.method = "first")

# Sort table by rank
predictor_summary <- predictor_summary[order(predictor_summary$Rank), ]

# Render table using knitr::kable
kable(
  predictor_summary,
  caption = "Table 8. Ranking of COVID-19 Death Predictors by Deviance",
  col.names = c("Predictor", "Deviance vs Diabetes", "Rank"),
  digits = 2,
  align = c("l", "r", "r")
)
```

Table 8 shows that when comparing the Hypertension and Diabetes model it is the only positive deviance, meaning the Hypertension model fits better than the Diabetes model and that Hypertension and Diabetes are the strongest single predictors of COVID-19 death in our dataset. Positive deviance indicates the second model is better. All other comparisons had Negative deviance values in LRT comparisons, indicating that the second model is worse than the first. For example, when comparing Diabetes and Smoking, the negative deviance indicates that Model 2 (Smoking) fits worse than Model 1 (Diabetes). Meaning that Diabetes is a stronger predictor than Smoking for COVID deaths.

# Discussion
## Hypothesis 1: Impact of Pre-existing Comorbidities on Mortality 
The results provide robust evidence supporting the hypothesis that pre-existing comorbidities significantly elevate the risk of mortality of COVID-19 patients in Mexico City. Univariate logistic regression analysis (Table 5) confirms that all four examined conditions—Diabetes, Hypertension, Obesity, and Smoking—are statistically significant predictors of mortality (p<0.001). The Odds Ratios (OR) for these conditions range from 1.54 to 1.62, indicating that patients with any one of these comorbidities have approximately 50-60% higher odds of mortality compared to those without. This association is visually reinforced by the heatmap in Figure 2, where the intense red coloration across the diagonal confirms that the probability of these associations occurring by chance is virtually zero.

## Hypothesis 2: Hypertension as the Leading Risk Factor
The data supports the hypothesis that Hypertension is the most critical single risk factor among the four comorbidities identified. While Diabetes shares the highest Odds Ratio (1.62) with Hypertension in single-predictor models, Hypertension exhibits the strongest statistical signal, evidenced by the highest z-statistic (19.78) and the most significant p-value (4.38×10 −87). Furthermore, model selection metrics in Table 6 identify the Hypertension-only model as having the lowest Akaike Information Criterion (AIC = 65,453.78) and Bayesian Information Criterion (BIC), suggesting it provides the best trade-off between model fit and complexity. This dominance is further corroborated by Figure 3, where Hypertension displays the highest Average Incremental Value (AIV), surpassing Obesity, Diabetes, and Smoking in its contribution to predicting mortality.

## Hypothesis 3: Cumulative Risk of Hypertension Combined with Other Comorbidities
The analysis validates the hypothesis that the combination of Hypertension with other comorbidities, particularly Diabetes, represents a more severe prognostic profile. Table 4 demonstrates that the risk estimates for Hypertension and its paired comorbidities remain stable and independent (e.g., OR ~1.62 for both Hypertension and Diabetes in the paired model), indicating an additive risk rather than redundancy. The paired model analysis in Table 7 identifies the Diabetes + Hypertension combination as the superior statistical model, achieving the lowest AIC (65,131.99) among all paired configurations. This confirms that while Hypertension is a primary driver, its co-occurrence with Diabetes creates the most informative predictive model for mortality, implying that patients presenting with this specific dual morbidity warrant the highest level of clinical vigilance.


# Conclusion

This current project provides a comprehensive analysis about the impact of common comorbidities on COVID-19 mortality. Based on the results, it was proven that there is a significant role of comorbidities and pre-exisiting health conditions on the likelihood of death. Among the comorbidities, hypertension was identified as the single most critical risk factor, demonstrating both the highest statistical robustness and the greatest incremental value for mortality prediction.

While Diabetes and Obesity are both severe risk factors, their roles differ slightly in modeling contexts: Diabetes offers high statistical confidence, whereas Obesity offers superior incremental predictive gain. Smoking, while significant, presents a lower risk profile relative to these metabolic conditions.

These findings support a risk-stratification strategy that prioritizes patients with Hypertension, particularly when combined with Obesity or Diabetes, for early intervention and intensive monitoring of vulnerable sectors of the population. Ultimately, this current project recommends future research to focus on the mechanistic interactions between these comorbidities, specifically hypertension, to further refine prognostic models.


